---
title: "results-draft"
author: "Pedro Nascimento de Lima"
date: "2024-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

# Reproducibility notes

This `quarto` document produces the results presented in *Nascimento de Lima et al. (2024). The value of environmental surveillance for pandemic response*. For details on how to install dependencies, see the `README.md` file. Before attempting to run this document, install all dependencies and run the `run_analysis.R` script. You should be able to reproduce all figures and tables from our paper using this document on a modern laptop within 30 minutes.

# Running the model

The following code block uses results stored in the `./output/r.rds` file or runs the script that generates this file.

```{r}
# The r object contains the results objects used in this file
if (!file.exists("./output/r.rds")) {
  source("./R/scripts/run_analysis.R")
} else {
  source("./R/library.R")
  r <- readRDS("./output/r.rds")
}
```

# Scenarios
Supplementary Table 1 provides details on parameter values used under our baseline scenario. The baseline scenario with conventional surveillance assumes a syndromic surveillance system based upon confirmed case counts, and there is a thirteen-day time span from exposure to the pathogen to detection by the public health authority. This thirteen-day lag covers the disease incubation phase (6.6 days)(cite) and testing and reporting delays (6 days).(cite) Consistent with estimates from the COVID-19 pandemic, the conventional surveillance system has a 30% case ascertainment rate.(cite) We assume the environmental sampling system can provide absolute, unbiased disease incidence estimates that lead the case counts by five days in the base-case scenario. We explore longer and shorter lead times, as this lag depends on two to ten days. We use 5 days as the base-case scenario, as wastewater surveillance systems have been able to demonstrate. (cite) Moreover, this analysis assumed that all the populations within the jurisdictions are covered by the EWS (or equivalently, that the sampled population is a representative sample of all the population).

# Population
The model simulates disease percolating through six interconnected jurisdictions that independently make nonpharmaceutical decisions. Individuals can travel and work across jurisdiction lines, and the six jurisdictions each have a population of 10^6 individuals. Mortality parameters were chosen to reflect mortality rates observed in the United States during the first year of the COVID-19 pandemic; hence our results can be interpreted as expected outcomes for a cluster of US counties.

# Outcomes
The average mortality per 100,000 population across the jurisdiction is the primary health outcome in this study. We also compute the cost of illness due to disease, using SARS-Cov-2 as the base-case scenario. We also account for economic costs caused by the nonpharmaceutical interventions introduced (discussed in the Supplementary Appendix). All outcomes are computed over a one-year time frame from the pandemic onset and are averages computed across the jurisdictions in our model. Since the model is stochastic, we simulate the model 1,000 times and present means and 95% prediction intervals (computed with the 2.5% and the 97.5% percentiles) of each outcome.

# Calibration: Add as a footnote to tau.
We use parameter estimates from the literature whenever possible (Supplementary Table 1). Parameters that are unobserved in the data, notably tau and. R0 is computed based on disease progression rates. We calibrate only the $tau$ parameter, which is highly heterogeneous and context-dependent (cite Laura), to match the number of deaths observed during the first year in the United States. We find that $tau = 0.137$. We explore the implication of a wider range of policy effectiveness in sensitivity analyses and explore this particular parameter in more detail.

## Sensitivity Analyses
We explore the performance of the EWS under conditions that depart from COVID-19 by varying model parameters one at a time, aiming to represent a new emerging pathogen against which the population has no immunity. Since a probabilistic characterization of uncertainty to parameters is elusive, we refrain from making probabilistic statements 

# Results

## Early warning systems can shape disease and policy dynamics

**MAybe this is useful for a briefing or slides, but maybe not the paper**:
An early warning system that shifts the epidemiological time-series used to inform non-pharmaceutical decisions can change the course of disease when used with effective interventions (Figure 1). We consider a network of jurisdictions that are independently managed by their own public health departments. For simplicity, we consider two cases: One in which the jurisdictions base their decisions upon syndromic surveillance data, and another (counterfactual) scenario where they are able to use data from environmental sampling sources (i.e., without loss of generality, wastewater surveillance), which lead the conventional surveillance data by a few days. The two sets of jurisdictions are otherwise equal, and face the exact same pandemic. 

## Value of EWS under baseline assumptions

<!--EWS systems provide meaningful value under base-case assumptions representing a hypothetical new pandemic with parameters mirroring those of SARS-Cov-2. -->

Assuming no NPIs interventions are put in place (No NPIs scenario), a new COVID-19-like pandemic would produce significant damage, causing `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "deaths_per_100k")` deaths per 100,000 people [mean (95% prediction interval of stochastic replications)] over a one-year time-frame (Table 1). In the absence of any behavioral change, the unmitigated pandemic would infect `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "epi_size")` percent of the population, causing `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "CH_illness")` dollars per person due to illness, and `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "CH_deaths")` dollars due to deaths. The total health cost would be projected at `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "CH")` dollars per person.

If policymakers followed the base-case NPI policy simulated in this paper, the death toll of the pandemic would be reduced from `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "deaths_per_100k")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "deaths_per_100k")` deaths per 100,000 population if no enhanced early-warning system (EWS) is adopted (NPIs w/o EWS scenario). As it was the case during the COVID-19 pandemic, such an achievement would involve a substantial public health effort. Under baseline assumptions, this effort would require `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "L1plus_days")` days of public health interventions, of which `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "L5_days")` days are spent under the most-restrictive policy intervention level. We project this effort to cost on the order of `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "CNPI")` per person. Based on those assumptions, the total cost of the pandemic would be reduced from `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "C")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "C")` per person.

Under baseline assumptions, the EWS provides net-positive value to society because they enhance the effectiveness of NPIs (i.e., preventing illness and death) more than they increase the costs of additional interventions introduced due to the early warning. Even an early warning system that provides only a 2-day early warning relative to a conventional pandemic response system (NPI + 2-day EWS scenario) is projected to reduce the pandemic death toll from `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "deaths_per_100k")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs + 2-day EWS", variable == "deaths_per_100k")` deaths per 100,000 population. Health benefits increase if the system can provide earlier warning. A system that provides a 5-day early warning (NPIs + 5-day EWS scenario) would be projected to result in `r pull_estimate(r$table_1_long, Scenario == "NPIs + 5-day EWS", variable == "deaths_per_100k")` deaths per 100,000 population, and a 10-day early warning would result in `r pull_estimate(r$table_1_long, Scenario == "NPIs + 10-day EWS", variable == "deaths_per_100k")` deaths per 100,000 people.

The use of early warning would result in earlier public health intervention, increasing the number of days of under NPIs from `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "L1plus_days")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs + 5-day EWS", variable == "L1plus_days")` days for a system providing a 5-day early warning signal. That said, the number of days at the *maximum* intervention level (i.e., a lockdown) was projected to *decrease* with the early warning system, as earlier action prevented surges from reaching alarming levels that prompted blunt interventions. For example, a system that provided a 5-day early warning reduced the number of days under the maximum intervention level from `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "L5_days")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs + 5-day EWS", variable == "L5_days")` days. Still, the net effect of the EWS was to marginally increase NPI intervention costs, from `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "CNPI")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs + 5-day EWS", variable == "CNPI")` dollars per person, assuming a 5-day early warning. Health and economic outcomes weighted, the early warning system reduced the total pandemic costs from `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "C")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs + 5-day EWS", variable == "C")` dollars per person. The difference between the two costs is the net-monetary benefit of the system, which is estimated at `r pull_estimate(r$table_1_long, Scenario == "NPIs + 5-day EWS", variable == "NMB")` dollars per person in the first year of the pandemic.

<!--
Points from this section for discussion:
* **EWS provides meaningful value under base-case assumptions.** Net-monetary benefit from the surveillance system is on the order of 1,000 dollars per person over a one-year time-frame. This outweighs any available estimate of costs.
* **Base-case results align with observed COVID-19 data.** US obseved on the order of 160 deaths per 100,000 people during the first year of the pandemic, and many states.
-->

## Surveillance net benefit depends on epidemiological and societal parameters

### Epidemiological parameters

A pathogen that is 50% more transmissible than wild-type SARS-Cov-2 would pose a greater challenge for pandemic response, causing `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "deaths_per_100k", NMB_comparator == T)` deaths per 100,000 people even if mitigated with the same NPI policy considered in our base-case scenario (Supplementary Table 2). Such a pandemic would cause a substantial societal cost of `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "C", NMB_comparator == T)` dollars per person. In those conditions, the EWS would be particularly valuable, reducing the death toll of the pandemic to `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "deaths_per_100k", NMB_comparator == F)` per 100,000 people, and would be projected to provide an NMB of `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "NMB", NMB_comparator == F)` dollars per person. Conversely, a pathogen that is 50% less transmissible than SARS-Cov-2 would result in a lower death toll of `r pull_estimate(r$table_1_long_all, Scenario == "0.5x transmissible", variable == "deaths_per_100k", NMB_comparator == T)` without the EWS system, reducing the net-monetary benefit of the system to `r pull_estimate(r$table_1_long_all, Scenario == "0.5x transmissible", variable == "NMB", NMB_comparator == F)` dollars per person. The value of the EWS system also depends on the infection fatality rate of the pathogen, but the net benefit of the EWS would be positive even for a pathogen that is half as deadly as SARS-Cov-2. All else equal, the EWS would still yield an NMB of `r pull_estimate(r$table_1_long_all, Scenario == "0.5x deadly", variable == "NMB", NMB_comparator == F)` dollars per person in a pandemic caused by a pathogen that is 50% less deadly than SARS-Cov-2, and would yield `r pull_estimate(r$table_1_long_all, Scenario == "1.5x deadly", variable == "NMB", NMB_comparator == F)` dollars per person in responding to a pathogen that is 50% more deadly than COVID-19.

### NPI Costs

NPI costs had a moderate effect on the value of surveillance systems. If NPI costs were 50% higher than assumed in our base-case scenario (i.e., if achieving the same NPI effectiveness were to cost 50% more than assumed under our base-case scenario), then the total cost of the pandemic would increase from `r pull_estimate(r$table_1_long_all, Scenario == "Base-case", variable == "C", NMB_comparator == T)` to `r pull_estimate(r$table_1_long_all, Scenario == "1.5x NPI costs", variable == "C", NMB_comparator == T)` dollars per person with the conventional surveillance system. (Supplementary Figure 1, **Supplementary Table X**). Despite that increase, the NMB of the system would be only marginally reduced from `r pull_estimate(r$table_1_long_all, Scenario == "Base-case", variable == "NMB", NMB_comparator == F)` to `r pull_estimate(r$table_1_long_all, Scenario == "1.5x NPI costs", variable == "NMB", NMB_comparator == F)` dollars per person, because the EWS system only marginally increased the number of days under interventions.

### Policy lags and stringency

Longer decision and implementation lags have an overall negative effect on the total costs of the pandemic, longer decision lags also would *increase* the net benefit of EWS systems for pandemic response. Our base-case scenario assumes that decision-makers take one week to decide and implement non-pharmaceutical interventions. If policymakers took two weeks to introduce and implement interventions, the total pandemic cost increases from `r pull_estimate(r$table_1_long_all, Scenario == "Base-case", variable == "C", NMB_comparator == T)` to `r pull_estimate(r$table_1_long_all, Scenario == "2-week decision lag", variable == "C", NMB_comparator == T)` dollars per person, assuming conventional surveillance. The early warning system effectively helps to roll-back decision lags, reducing total costs to `r pull_estimate(r$table_1_long_all, Scenario == "2-week decision lag", variable == "C", NMB_comparator == F)` dollars per person, yielding an NMB of `r pull_estimate(r$table_1_long_all, Scenario == "2-week decision lag", variable == "NMB", NMB_comparator == F)` relative to conventional surveillance. Conversely, in contexts where NPI implementation is swift (i.e, decisions are made within 3 days), the NMB of EWS would be lower but still positive at `r pull_estimate(r$table_1_long_all, Scenario == "3-day decision lag", variable == "NMB", NMB_comparator == F)` dollars per person.

Policy stringency (i.e., how stringent the case thresholds used to introduce or remove interventions) has a similar effect on the value of EWS relative to decision lags - the EWS provides the most value under the least desirable conditions for disease prevention. In a context where decision-makers are 50% more stringent than in our base-case assumptions (i.e., their case thresholds are half of its base-case), society is able to avert a substantial number of deaths, reducing mortality from `r pull_estimate(r$table_1_long_all, Scenario == "Base-case", variable == "deaths_per_100k", NMB_comparator == T)` to `r pull_estimate(r$table_1_long_all, Scenario == "1.5x stringent", variable == "deaths_per_100k", NMB_comparator == T)` deaths per 100,000 population without the EWS. In that scenario, the EWS yields `r pull_estimate(r$table_1_long_all, Scenario == "1.5x stringent", variable == "NMB", NMB_comparator == F)` dollars per person of net benefit since most of the deaths would already be prevented through more stringent NPIs. Conversely, if policymakers were more lenient policy choices would change the system overall outcomes but also imply a higher value of the enhanced surveillance system.

### NPI effectiveness

NPI effectiveness (i.e., the % reduction in infectious contacts caused by NPIs) had a non-monotonic effect on the value of surveillance systems to pandemic response (**Supplementary Figure X**). If NPI effectiveness is 0%, an improved surveillance system has no value, as expected. The value of the enhanced surveillance system increases with NPI effectiveness slowly for maximum NPI effectiveness below around 50%, and peaks when the maximum NPI effectiveness reaches 60%, enough to reduce the baseline R0 of 2.5 below 1 at the pandemic onset. In general, we find that the value of the EWS peaks at a maximum NPI effectiveness value at $1-1/R0$, for different values of R0. Beyond that point, increases in NPI effectiveness result in lower total pandemic costs, but the value of the surveillance system starts to decrease.
